### 2017-07-12 pkg/model4you/inst/JORS/man_model4you.Rnw
more text
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1500 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass{jors}

@@

\section*{Abstract}
Typical models estimating treatment effecs assume that the treatment effect is
the same for all individuals observed and for possible future individuals.
Model-based recursive partitioning allows to relax this assumption.  With
model-based trees one can compute treatment effects for different strata of
individuals. The strata are found in a data driven fashion and depend on
characteristics of the individuals.  Model-based random forests allow for a
similarity estimation between individuals in terms of model estimates and
through this we can estimate personalised models.  The R package
\emph{model4you} (available on CRAN) implements these stratified and
personalised models with a focus on ease of use and interpretability so that
clinicians and other users can take the model they usually use for the
estimation of the average treatment effect and with a few lines of code get a
visualisation that is easy to understand and interpret.

@@

%\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles. A short comparison with software which implements similar functionality should be included in this section. }
Studies in various fields randomly assign individuals one of two groups with
different exposure and then measure a response. For example, in clinical trials
patients are assigned to one of two treatment groups where usually one
treatment group receives a new treatment or drug and the other treatment group
receives the standard of care or a placebo.

@@

If you do not believe that all individuals have the same intercept and
treatment effect the R package \emph{model4you} gives you two options to loosen
up this strict model: \paragraph{(1) Model-based trees} identify subgroups
where within the subgroups the model parameters are similar and between groups
the model parameters are different. This is achieved by finding instabilities
in the model parameters with respect to a variable (characteristicum) and
recursively partitioning the data into groups. If, for example the algorithm
finds that men and women have differing treatment effects, the data is
partitioned into two subgroups. Details on model-based trees in general can be
found in \citep{zeileis_model-based_2008} and for the special use case in
clinical trial in \citep{seibold_model-based_2016}.

Just a single line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.

In R, to get from the initial base model to personalised models for all
observations in the training data, type:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

\begin{figure}
\centering
\include{pmworkflow}
\caption{Personalised models workflow: from real world to interpretable
visualisation.}
\end{figure}

In the following we will present two example applications for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for plots we will need the
\emph{ggplot2} package and we use a black and white theme.
<<prep>>=
library("model4you")
set.seed(2017)

library("ggplot2")
theme_set(theme_classic())
@

\paragraph{Mathematics exam analysis}
<<math_data>>=
data("MathExam14W", package = "psychotools")

@@

<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
## compute base model
bmod_math <- lm(pcorrect ~ group, data = MathExam)
lmplot(bmod_math, densest = TRUE)
@

@@

\paragraph{Analysis of the German Breast Cancer Study Group 2}

<<GBSG2_prep>>=
library("survival")
data("GBSG2", package = "TH.data")
@

<<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
survplot(bmod)

## extract coefficients and confidence intervals
coefs <- coef(bmod)
ci <- confint(bmod)
coefs <- cbind(coefs, ci)

## prepare for plotting
digits <- 2
cf <- format(round(coefs, digits), nsmall = digits)
colnams <- colnames(cf)
colnams[1] <- "theta"
grid.table(cf, cols = colnams,
           theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
           vp = viewport(x = 0, y = 0, 
                         width = 0.7, height = 0.4,
                         just = c("left", "bottom")))
@

<<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
tr <- pmtree(bmod)
plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
                                          confint = TRUE))
@

<<GBSG2_forest>>=
frst <- pmforest(bmod)
@

<<GBSG_pmod1>>=
newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
                           pnodes = 3, progrec = 50, estrec = 60)
mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
@

<<GBSG2_pmod>>=
coeffun <- function(model) {
  ## model coefficients
  coefs <- c(coef(model), scale = model$scale)

  ## difference in median survival 
  p = 0.5
  coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
                                 scale = exp(coefs["(Intercept)"]))
  coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
                                 scale = exp(coefs["(Intercept)"] + coefs[2]))
  coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]

  return(coefs)
}
coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)

## add to data and plot
dpdat <- cbind(coefs, GBSG2)
znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]

dplot <- function(z) {
  p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
  if(is.factor(dpdat[, z])) {
    p + geom_boxplot()
  } else {
    p + geom_point(alpha = 0.2) + 
      geom_smooth(fill = NA, method = "loess")
  }
}
dps <- lapply(znam, dplot)
do.call(grid.arrange, dps)

do.call(grid.arrange, lapply(dps, function(p) {
  dmin <- min(dpdat$median_sdiff)
  dq <- quantile(dpdat$median_sdiff, 0.99)
  p + coord_cartesian(ylim = c(dmin, dq))
}))
@

\section*{Implementation and architecture}

Based on partykit, merely convenience interface.

\begin{figure}
\centering
\include{treefamily}
\caption{The family of model-based trees and forests.}
\end{figure}

>>>>

\documentclass{josr}

@@

\section*{Abstract}
Typical models estimating treatment effects assume that the treatment effect is
the same for all individuals.
Model-based recursive partitioning allows to relax this assumption and allows for
the estimation of stratified treatment effects (model-based trees) or even
personalised treatment effects (model-based forests).  With
model-based trees one can compute treatment effects for different strata of
individuals. The strata are found in a data driven fashion and depend on
characteristics of the individuals.  Model-based random forests allow for a
similarity estimation between individuals in terms of model estimates. The similarity
measure can then be used to estimate personalised models. 
The R package \emph{model4you} implements these stratified and
personalised models with a focus on ease of use and interpretability so that
clinicians and other users can take the model they usually use for the
estimation of the average treatment effect and with a few lines of code get a
visualisation that is easy to understand and interpret.

@@

%\textcolor{blue}{An overview of the software, how it was produced, and the research for which it has been used, including references to relevant research articles. A short comparison with software which implements similar functionality should be included in this section. }
Studies in various fields randomly assign individuals to one of two groups with
different exposure and then measure a response. For example, in clinical trials
patients are assigned to one of two treatment groups where usually one
treatment group receives a new treatment or drug and the other treatment group
receives the standard of care or a placebo.

@@

For cases where the assumption that all individuals have the same intercept
and treatment effect is too strict the R package \emph{model4you} offers two 
options:

\paragraph{(1) Model-based trees} identify subgroups
where within the subgroups the model parameters are similar and between groups
the model parameters are different. This is achieved by finding instabilities
in the model parameters with respect to a variable (characteristicum) and
recursively partitioning the data into groups. If, for example the algorithm
finds that men and women have differing treatment effects, the data is
partitioned into two subgroups. Details on model-based trees in general can be
found in \citep{zeileis_model-based_2008} and for the special use case in
clinical trial in \citep{seibold_model-based_2016}.
Just a single line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@

\paragraph{(2) Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.
To compute personalised models for all observations in the training data is 
simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@

% \begin{figure}
% \centering
% \include{pmworkflow}
% \caption{Personalised models workflow: from real world to interpretable
% visualisation.}
% \end{figure}

In the following we will present two example applications for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for plots we will need packages
\emph{ggplot2} and \emph{gridExtra}.
<<prep>>=
library("model4you")
set.seed(2017)

library("ggplot2")
theme_set(theme_classic())
library("gridExtra")
@

\paragraph{Mathematics exam analysis}
In 2014 first-year business and economics students at the University of Innsbruck
were divided into two examination groups. Group 1 wrote the exam in the morning and
group 2 started after the first group finished.

<<math_data>>=
data("MathExam14W", package = "psychotools")

@@

<<math_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for Mathematics Exam data.">>=  
## compute base model
bmod_math <- lm(pcorrect ~ group, data = MathExam)
lmplot(bmod_math, densest = TRUE)

## extract coefficients and confidence intervals
coefs <- coef(bmod_math)
ci <- confint(bmod_math)
coefs <- cbind(coefs, ci)

## prepare for plotting
digits <- 2
cf <- format(round(coefs, digits), nsmall = digits)
colnams <- colnames(cf)
colnams[1] <- "theta"
gridExtra::grid.table(
  cf, cols = colnams,
  theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
  vp = viewport(x = 0.5, y = 0, width = 0.7, height = 0.4,
                just = c("center", "bottom"))
)
@

@@

% \paragraph{Analysis of the German Breast Cancer Study Group 2}
% 
% <<GBSG2_prep>>=
% library("survival")
% data("GBSG2", package = "TH.data")
% @
% 
% <<GBSG2_bmod, fig.height = 5, out.width = '0.7\\textwidth', fig.cap = "Base model for GBSG2 data.">>=
% bmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, model = TRUE)
% survplot(bmod)
% 
% ## extract coefficients and confidence intervals
% coefs <- coef(bmod)
% ci <- confint(bmod)
% coefs <- cbind(coefs, ci)
% 
% ## prepare for plotting
% digits <- 2
% cf <- format(round(coefs, digits), nsmall = digits)
% colnams <- colnames(cf)
% colnams[1] <- "theta"
% gridExtra::grid.table(
%   cf, cols = colnams,
%   theme = ttheme_minimal(colhead = list(fg_params = list(parse = TRUE))),
%   vp = viewport(x = 0, y = 0, width = 0.7, height = 0.4,
%                 just = c("left", "bottom"))
% )
% @
% 
% <<GBSG2_tree, fig.width = 9, fig.cap = "Personalised model tree for GBSG2 data.">>=
% tr <- pmtree(bmod)
% plot(tr, terminal_panel = node_pmterminal(tr, plotfun = survplot, 
%                                           confint = TRUE))
% @
% 
% <<GBSG2_forest>>=
% frst <- pmforest(bmod)
% @
% 
% <<GBSG_pmod1>>=
% newpatient_i <- data.frame(age = 60, menostat = "Post", tsize = 20, tgrade = "II",
%                            pnodes = 3, progrec = 50, estrec = 60)
% mod_i <- pmodel(x = frst, fun = identity, newdata = newpatient_i)
% @
% 
% <<GBSG2_pmod>>=
% coeffun <- function(model) {
%   ## model coefficients
%   coefs <- c(coef(model), scale = model$scale)
%   
%   ## difference in median survival 
%   p = 0.5
%   coefs["median_s0"] <- qweibull(p = p, shape = 1/coefs["scale"], 
%                                  scale = exp(coefs["(Intercept)"]))
%   coefs["median_s1"] <- qweibull(p = p, shape = 1/coefs["scale"], 
%                                  scale = exp(coefs["(Intercept)"] + coefs[2]))
%   coefs["median_sdiff"] <- coefs["median_s1"] - coefs["median_s0"]
%   
%   return(coefs)
% }
% coefs <- pmodel(x = frst, fun = coeffun, OOB = TRUE)
% 
% ## add to data and plot
% dpdat <- cbind(coefs, GBSG2)
% znam <- names(GBSG2)[!(names(GBSG2) %in% c("horTh", "time", "cens"))]
% 
% dplot <- function(z) {
%   p <- ggplot(dpdat, aes_string(x = z, y = "median_sdiff")) 
%   if(is.factor(dpdat[, z])) {
%     p + geom_boxplot()
%   } else {
%     p + geom_point(alpha = 0.2) + 
%       geom_smooth(fill = NA, method = "loess")
%   }
% }
% dps <- lapply(znam, dplot)
% do.call(grid.arrange, dps)
% 
% do.call(grid.arrange, lapply(dps, function(p) {
%   dmin <- min(dpdat$median_sdiff)
%   dq <- quantile(dpdat$median_sdiff, 0.99)
%   p + coord_cartesian(ylim = c(dmin, dq))
% }))
% @

\section*{Implementation and architecture}
\begin{figure}
\centering
\include{treefamily}
\caption{The family of model-based trees and forests. Square nodes represent
functions, round nodes represent function arguments. Solid lines represent 
direct usage, dashed lines indirect usage (iteration between tree and model 
estimation), and dotted lines represent close similarity but no dependency.}
\label{fig:family}
\end{figure}

The R package \emph{model4you} is based on the \emph{partykit} package. In fact, 
everything that can be done with \emph{model4you} can already be done with 
\emph{partykit} (version $\geq$ 1.2-0). \emph{model4you} is merely a convenience 
interface and focusses on ease of use and interpretability. Various other packages
depend on \emph{partykit}. Figure~\ref{fig:family} shows all packages depending
on \emph{partykit} that deal with model-based trees and forests. Functions 
\Sexpr{'partykit::ctree()'} and \Sexpr{'partykit::mob()'} provide the basis for all
others. Note that \Sexpr{'partykit::ctree()'} computes conditional inference trees
but by specifying argument \Sexpr{'ytrafo'} appropriately, it can also compute
model-based trees. This functionality is used for packages \emph{disttree} 
(partitioning of Generalized Additive Models for Location, Scale and Shape),
\emph{trtf} (partitioning of transformation models) and \emph{model4you}. 
Many of the packages shown in Figure~\ref{fig:family} are available on CRAN. 
The remaining packages are under development and available on 
\url{https://r-forge.r-project.org/projects/partykit/}.
